\thispagestyle{empty}
{\Huge\forallx}

{\tt An Introduction to Formal Logic}

\vfill


{\sf P.D. Magnus}\\
\emph{University at Albany, State University of New York}


\vfill


{\sf
	fecundity.com/logic, version 1.31 [\bookversion]\\
	This book is offered under a Creative Commons license.\\
	(Attribution 4.0)
}



\newpage
\thispagestyle{empty}%
{\sf
The author would like to thank the people who made this project possible. Notable among these are Cristyn Magnus, who read many early drafts; Aaron Schiller, who was an early adopter and provided considerable, helpful feedback; {and} Bin Kang, Craig Erb, Nathan Carter, Wes McMichael, Selva Samuel,  Dave Krueger, Brandon Lee, Toan Tran, Marcus Adams, Matthew Brown, and the students of Introduction to Logic, who detected various errors in previous versions of the book.
}

\vfill
{
\copyright\ \ifthenelse{\year=2005}{\number\year}{2005--\number\year} by P.D. Magnus. Some rights reserved.
}

{\footnotesize
You are free to copy this book, to distribute it, to display it, and to make derivative works, under the following condition: Attribution. You must give the original author credit. --- For any reuse or distribution, you must make clear to others the license terms of this work. Any of these conditions can be waived if you get permission from the copyright holder. Your fair use and other rights are in no way affected by the above. --- This is a human-readable summary of the full license, which is available on-line at \url{http://creativecommons.org/licenses/by/4.0/}
}

{
Typesetting was carried out entirely in \LaTeX$2\varepsilon$. The style for typesetting proofs is based on fitch.sty (v0.4) by Peter Selinger, University of Ottawa.

This copy of \forallx\ is current as of \today. The most recent version is available on-line at \url{http://www.fecundity.com/logic}
}

